Numerical Investigation of a Neural Field Model Including Dendritic Processing

### 2020-03-24 Manuscript/manuscript.tex
First committ

\documentclass[a4paper]{siamart190516}

@@

% Sets running headers as well as PDF title and authors
\headers{Preliminary notes}{D. Avitabile}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{Preliminary notes}

% Authors: full names plus addresses.
\author{%
  Daniele Avitabile%
  \thanks{%
    Vrije Universiteit Amsterdam,
    Department of Mathematics,
    Faculteit der Exacte Wetenschappen,
    De Boelelaan 1081a,
    1081 HV Amsterdam, The Netherlands.
  \protect\\
    Inria Sophia Antipolis M\'editerran\'ee Research Centre,
    MathNeuro Team,
    2004 route des Lucioles-Boîte Postale 93 06902,
    Sophia Antipolis, Cedex, France.
  \protect\\
    (\email{d.avitabile@vu.nl}, \url{writemywebpage}).
  }
  % \and
%   Paul T. Frank \thanks{Department of Applied Mathematics, Fictional University, Boise, ID 
% (\email{ptfrank@fictional.edu}, \email{jesmith@fictional.edu}).}
% \and Jane E. Smith\footnotemark[3]
}

@@

\begin{abstract}
  Text to be inserted in the abstract
\end{abstract}

\section{Conventions} \label{sec:conventions}
See \texttt{docsiamart.pdf} for many examples. I generally adopt the conventions of
that template. I remark a few things below.

\subsection{References} \label{sec:references}
These are random citations \cite{KoMa14,siam}. For cross-references I use
\texttt{cleveref}: like this
\cref{eq:test}, like this \Cref{eq:test}, or like this \cref{sec:conventions}, or
this \cref{hyp:trivial}, or this \cref{clm:trivial}.
\begin{equation}\label{eq:test}
  x = y
\end{equation}

\subsection{Todos}
You can find templates of other author's todos in
\texttt{damacros.sty}, which you are welcome to edit.
This is how I make a short comment\da{Short comment by Daniele}. Comments can also go
inline like the one below.
\da[inline]{Long comments by Daniele} 
% \missingfigure
% 
% \section*{Acknowledgments}

\section{Testing commands}

\subsection{Testing maths operators}
Here are tests for operators
\[
  \spec, \quad 
  \range, \quad 
  \spn, \quad
  \diag, \quad
  \meas, \quad
  \id, \quad
  \real z, \quad
  \imag z, \quad
  \e, \quad
  \iunit, \quad
\]

\subsection{Testing additional theorem-like environments} The following additional
theorem-like environments may be useful.

\begin{hypothesis} \label{hyp:trivial}
  This is a hypothesis
\end{hypothesis}

\begin{claim} \label{clm:trivial}
  This is a claim
\end{claim}

\subsection{Testing sets}
This are a few examples of sets (see more commands in \texttt{damacros.sty})
\[
  \RSet, \quad
  \CSet, \quad
  \ZSet, \quad
  \sph, \quad
  \ldots
\]

\subsection{Empty arguments}
Command for empty arguments, which require spacing
\[
  u(\blank,t) \quad t \in \RSet
\]

\subsection{Calligraphic fonts}
If one really must use calligraphic fonts, there are macros for this.
\[
  \calA, \quad
  \calL, \quad
  \calM, \quad
  \calS, \quad
\]

>>>>

\documentclass[a4paper,final]{siamart190516}

@@

% Sets running headers as well as PDF title and authors
\headers{Neural fields with dendrites }{D. Avitabile, S. Coombes, P.~M. Lima}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{Numerical Investigation of a Neural Field Model Including Dendritic Processing}

% Authors: full names plus addresses.
\author{%
  Daniele Avitabile%
  \thanks{%
    Vrije Universiteit Amsterdam,
    Department of Mathematics,
    Faculteit der Exacte Wetenschappen,
    De Boelelaan 1081a,
    1081 HV Amsterdam, The Netherlands.
  \protect\\
    Inria Sophia Antipolis M\'editerran\'ee Research Centre,
    MathNeuro Team,
    2004 route des Lucioles-Boîte Postale 93 06902,
    Sophia Antipolis, Cedex, France.
  \protect\\
    (\email{d.avitabile@vu.nl}, \url{writemywebpage}).
  }
  \and
  Stephen Coombes \thanks{Centre for Mathematical Medicine and Biology, School of
  Mathematical Sciences, University of Nottingham, NG7 2RD, UK.}
  \and
  Pedro M. Lima \thanks{CEMAT, Instituto Superior Tecnico,University of Lisbon,
  Portugal}
%   Paul T. Frank \thanks{Department of Applied Mathematics, Fictional University, Boise, ID 
% (\email{ptfrank@fictional.edu}, \email{jesmith@fictional.edu}).}
% \and Jane E. Smith\footnotemark[3]
}

\usepackage[ruled,linesnumbered]{algorithm2e}

\newcommand{\setan}{{}_{\stackrel{\textstyle \longrightarrow}{n}}}

\newcommand{\Real}{\operatorname{Re}}    % real line
\newcommand{\Imag}{\operatorname{Im}}

\newcommand{\Cb}{\mathbb{C}}
\newcommand{\Rb}{\mathbb{R}}
\newcommand{\Nb}{\mathbb{N}}
\newcommand{\Zb}{\mathbb{Z}}
\newcommand{\Ib}{\mathbb{I}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\zast}{\stackrel{\ast}{z}}
\newcommand{\xast}{\stackrel{\ast}{x}}
\newcommand{\yast}{\stackrel{\ast}{y}}
\newcommand{\xastum}{\stackrel{\ast}{x_1}}
\newcommand{\xastdois}{\stackrel{\ast}{x_2}}

\newcommand{\limz}{\lim_{z \rightarrow \zast}} 
\newcommand{\diff}{\mathop{}\!\mathrm{d}}
\newcommand{\ep}{\varepsilon}

\graphicspath{ {Figures/} }

%\DeclareMathOperator{\e}{e}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\supp}{supp}

% Commands for tables
\usepackage{ctable}
\newcommand{\otoprule}{\midrule[\heavyrulewidth]}
\usepackage{color}
\usepackage{listings}

\definecolor{LightGrey}{rgb}{0.9629411,0.9629411,0.9629411}
\definecolor{LighterGrey}{gray}{0.99}
\definecolor{Mauve}{rgb}{0.58,0,0.82}
\definecolor{Emerald}{rgb}{0.31, 0.78, 0.47}
\definecolor{RoyalBlue}{rgb}{0.25, 0.41, 0.88}
\definecolor{myGreen}{cmyk}{0.82,0.11,1,0.25}

\lstset{
    language=matlab,
    morekeywords={ifftshift,refreshdata},
    keywordstyle=\color{RoyalBlue},
    basicstyle=\scriptsize\ttfamily,
    % commentstyle=\color{Emerald}\scriptsize\ttfamily,
    commentstyle=\color{myGreen}\scriptsize\ttfamily,
%     directivestyle=\color{Mauve}\scriptsize\ttfamily,
    showspaces=false,            
    showstringspaces=false,
    stringstyle=\color{Mauve}\scriptsize\ttfamily,
    numbers=none,
    numberstyle=\scriptsize,
    stepnumber=1,
    numbersep=8pt,
    showstringspaces=false,
    breaklines=true,
    frameround=ftff,
    frame=lines,
    backgroundcolor=\color{LightGrey}
} 
\def\inline{\lstinline[basicstyle=\ttfamily,keywordstyle={},directivestyle={}]}

@@

\begin{abstract}
We consider a simple neural field model in which the state variable is dendritic
voltage, and in which somas form a continuous one-dimensional layer.
This \textit{neural field} model with \textit{dendritic processing} is formulated as
an integro-differential equation. We introduce a
computational method for approximating solutions to this nonlocal model, and use
it to perform numerical simulations for neuro-biologically realistic choices of
anatomical connectivity and nonlinear firing rate function.  For the time
discretisation we adopt an Implicit-Explicit (IMEX) scheme; the space discretisation
is based on a finite-difference scheme to approximate the diffusion term and uses the
trapezoidal rule to approximate integrals describing the nonlocal interactions in
the model. We prove that the scheme is of first-order in time and second order in
space, and can be efficiently implemented if the factorisation of a small, banded
matrix is precomputed. By way of validation we compare the outputs of a numerical
realisation to theoretical predictions for the onset of a Turing pattern, and to the
speed and shape of a travelling front for a specific choice of Heaviside firing rate.
We find that theory and numerical simulations are in excellent agreement.

\end{abstract}

\section{Introduction}

Ever since Hans Berger made the first recording of the human electroencephalogram
(EEG) in 1924 there has been a tremendous interest in understanding the physiological
basis of brain rhythms. This has included the development of mathematical models of
cortical tissue, which are often referred to as neural field models.  
The formulation of these models has not changed much since the seminal work of Wilson
and Cowan, Nunez and Amari in the 1970s, as recently described in \cite{Coombes2014}.
Neural fields and neural mass models approximate neural activity assuming the
cortical tissue is a continuous medium. They are coarse-grained spatiotemporal models,
which lack important physiological mechanisms known to be fundamental in generating brain
rhythms, such as dendritic structure and cortical folding. Nonetheless their basic
structure has been shown to provide a mechanistic starting point for understanding
whole brain dynamics, as described by Nunez \cite{Nunez1995}, and especially that of
the EEG.

Modern biophysical theories assert that EEG signals from a single scalp
electrode arise from the coordinated activity of $\sim 10^8$ pyramidal cells in the
cortex.  These are arranged with their dendrites in parallel and perpendicular to the
cortical surface. When activated by synapses at the proximal
dendrites, extracellular current flows parallel to the dendrites, with a net
membrane current at the synapse. For excitatory (inhibitory) synapses this creates a
sink (source) with a negative (positive) extracellular potential.  Because there is
no accumulation of charge in the tissue the proximal synaptic current is compensated
by other currents flowing in the medium causing a distributed source in the case of a
sink and vice-versa for a synapse that acts as a source. Hence, at the population
level the potential field generated by a synchronously activated population of
cortical pyramidal cells behaves like that of a dipole layer.  Although the important
contribution that single dendritic trees make to generating extracellular electric
field potentials has been known for some time, and can be calculated using
Maxwell's equations \cite{Pettersen08}, they are often not accounted for in neural
field models.  However, with the advent of laminar electrodes to record from
different cortical layers it is now timely to build on early work by Bressloff,
reviewed in \cite{Bressloff97}, and develop neural field models that incorporate a
notion of dendritic depth.  This will allow a significant and important departure
from present-day neural field models, and recognise the contribution of dendritic
processing to macroscopic large-scale brain signals. A simple way to generalise
standard neural field models is to consider the dendritic cable model of Rall as the
core component in a neural field, with source terms on the cable mediating
long-range synaptic interactions.  These in turn can be described with the
introduction of an \textit{axo-dendritic} connectivity function.

Here we consider a neural field model which treats the voltage on a dendrite as the
primary variable of interest in a simple model of neural tissue. The model comprises
a continuum of somas (a \emph{somatic layer}, see schematic in
Figure~\ref{fig:sketch}(a)). Dendrites are modeled as unbranched fibres,
orthogonal to the somatic layer which, for simplicity, is one-dimensional and
rectilinear (see Figure~\ref{fig:sketch}(b)). At each point along the somatic layer
$x \in \Rb$ we envisage a fibre with coordinate $\xi \in \Rb$. The voltage dynamics
along the fibre is described by the cable equation, with a nonlocal input current
arising as
an integral
over the outputs from the somatic layer (where $\xi=0$). Denoting the voltage by
$V(x,\xi,t)$ we have an integro-differential equation for the
real-valued function $V:  \Rb^2 \times \Rb  \rightarrow \Rb$ of the form
  \begin{multline} \label{1}
  \partial_t V(x,\xi, t) = (-\gamma + \nu \partial_{\xi \xi}) V(x,\xi,t)
                        + G(x,\xi,t)
			\\
  + \int_{\Rb^2} W(x,\xi,y,\eta) 
                        S(V(y,\eta,t))\diff y \diff \eta ,
  \end{multline}
posed on $(x,\xi,t) \in \Rb^3$, for some typically sigmoidal or
Heaviside-type  firing rate function $S$, and some external input function $G$. Here
$\nu$ is the diffusion coefficient and $1/\gamma$ the membrane time-constant of the
cable. As we shall see below, it will be crucial for our analysis that currents flow
exclusively along the fibres, that is, the diffusive term in \eqref{1} contains
derivatives only with respect to $\xi$.

The model is completed with a choice of the generalised axo-dendritic connectivity
function $W$. The nonlocal input current arises from the somatic layer, hence they
are transferred from sources in an $\ep$-neighbourhood of $\xi = 0$, $0 < \ep \ll 1$, to 
contact points in an $\ep$-neighbourhood of $\xi=\xi_0$ on the cable (see
Figure~\ref{fig:sketch}(b)). In addition, the strength of interaction depends
solely on the distance between the source and the contact point, measured along the
somatic layer, leading to the decomposition
\begin{equation}\label{eq:kernel}
  W(x,\xi,y, \eta)=    w(|x-y|)\delta_\ep(\xi-\xi_0) \delta_\ep(\eta),
\end{equation}
where $w$ describes the strength of interaction across the somatic space and is
chosen to be translationally invariant and $\delta_\ep$ is a quickly-decaying
function. 
% supported on a compact interval $I_\ep \subset \Rb$,
% of $O(\ep)$ measure, with $\ep \ll 1$.
%Gaussian function with standard deviation $\ep \ll 1$.

\begin{figure}
  \centering
  \includegraphics{sketch}
  \caption{Schematic of the neural field model. (a) Dendrites are represented as
    unbranched fibres (red), orthogonal to a continuum of somas (somatic layer, in
    grey). (b) Model
    \eqref{1} is for a 1D somatic layer, with coordinate
$x \in \Rb$, and fiber coordinate $\xi \in \Rb$. Input currents are generated in a
small neighbourhood of the somatic layer, at $\xi=0$ and are delivered to a contact
point, in a small neighbourhood of $\xi = \xi_0$. The strength of interaction depends
on the distance between sources and contact points, measured along the somatic layer,
hence the inputs that are generated at $A$ and transmitted to $B$, $C$, and $D$ depend
on $|x_B - x_A|$, $|x_C-x_A|$, and $|x_D - x_A|$, respectively (see
\eqref{eq:kernel}).}
  \label{fig:sketch}
\end{figure}

%%%%%  Introduce state of art, relationship with other neural field models, applications
This work introduces a computational method for approximating
solutions to (\ref{1}), subject to suitable initial and boundary conditions,
and applies it to the numerical simulation of the model with kernel given by
(\ref{eq:kernel}). Numerical methods for neural fields in 2-dimensional media have
been introduced recently in flat
geometries~\cite{Rankin2014,hutt2014numerical,lima2015numerical} and on 2-manifolds
embedded in a 3-dimensional space~\cite{Bojak2010,Visser:2017hy}. In addition,
several available open-source codes, such as the Neural Field
Simulator~\cite{Nichols2015}, the Brain Dynamics Toolbox~\cite{Heitmann}, and
NFTsim~\cite{SanzLeon2018}, perform simulations of neural field equations. Numerical
schemes for models of type~\eqref{1} have not been introduced, analysed, or
implemented, and these are the main contributions of the present article.

In Section \ref{Sec:numerical} we describe, analyse, and discuss implementation
details of the numerical method. In Sections \ref{sec:TWTest}--\ref{sec:TuringTest}
we illustrate the performance of the method by means of some numerical experiments,
including problems whose exact solution has known properties. The numerical results
are discussed and their physical meaning is explained. We finish with some
conclusions and discussion in Section \ref{sec:conclusions}.

\section{Numerical Scheme}
\label{Sec:numerical}
Numerical simulations are performed on \eqref{1}, posed on a
bounded, cylindrical somato-dendritic  domain 
\[
  \Omega = \Rb/2L_x\Zb \times (-L_\xi,L_\xi), %\qquad L_x,L_\xi \gg 1,
\]
%$\Omega = (-L_x,L_x) \times
%(-L_\xi,L_\xi) \cong
%\mathbb{S} \times (-L_\xi,L_\xi) $
and subject to initial and boundary conditions,
\begin{equation}\label{eq:systemNum}
  \begin{aligned}
    & \partial_t V  = (-\gamma + \nu \partial_{\xi \xi}) V + K(V) + G
    % & & \textrm{on $ (-L_x, L_x) \times (-L_\xi,L_\xi) \times (0,T]$}, \\
    & & \textrm{on $ \Omega \times (0,T]$}, \\
  &  V(\blank,\blank,0)  = V_0
    % & & \textrm{on $(-L_x, L_x) \times (-L_\xi,L_\xi)$}, \\
    & & \textrm{on $ \Omega$}, \\
  % &  \partial_x V(-L_x,\blank,\blank) = \partial_x V(L_x,\blank,\blank) = 0
  %   & & \textrm{on $ [-L_\xi,L_\xi] \times [0,T]$}, \\
  &  \partial_\xi V(\blank,-L_\xi,\blank) = \partial_\xi V(\blank,L_\xi,\blank) = 0
    & & \textrm{on $ (-L_x,L_x] \times [0,T]$}, \\
  \end{aligned}
\end{equation}
for some positive constants $T$, $L_x$, $L_\xi$. This setup implies
$2L_x$-periodicity in the somatic
direction, and Neumann boundary conditions in the dendritic direction.
We denote by $K$ the integral operator defined by
\[
  (K(V))(x,\xi,t) = \int_{\Omega} W_{\Omega}(x,\xi,y,\eta)
    S(V(y,\eta,t)) \diff y \diff \eta,
  \qquad (x,\xi) \in \Omega.
\]
where $W_{\Omega}$ is the restriction of $W$ on $\Omega$. This restriction implies
that the function $w$ in \eqref{eq:kernel} be substituted by
its periodic extension on $[-L_x,L_x)$. In the remainder of this paper we will omit the
subscript $\Omega$ from $W$, and assume $w$ to be $2L_x$-periodic.

To expose our scheme we introduce
a spatiotemporal discretisation 
%of the closed domain $\bar{\Omega} \times [0,T]$, 
using the evenly spaced grid 
$\{(x_j,\xi_i,t_n)\}$ defined by
\[
  \begin{aligned}
    & x_j = -L_x + j h_x, & & j \in \Nb_{n_x}, && h_x = 2L_x/n_x, \\
    & \xi_i   = -L_\xi + (i-1) h_\xi, & & i \in \Nb_{n_\xi}, && h_\xi =
    2L_\xi/(n_\xi-1), \\
    & t_n = n \tau, & & n \in \Nb_{n_t}, && \tau = T/n_t, \\
  \end{aligned}
\]
where we posed $\Nb_k = \{1,2,\ldots,k\}$ for $k \in \Nb$. The scheme we propose uses
the method of lines for \eqref{eq:systemNum}, in conjunction with differentiation
matrices for the diffusive term and a quadrature scheme for the integral operator.

% In the description of the numerical scheme, it will be useful to alternate between
% a matricial representation of the approximate values of $V(x,\xi,t)$ at the
% gridpoints, which will be denoted with a slight abuse
% of notation as
% \[
%   V(t) = \{ V_{ij}(t) \colon (i,j) \in \Nb_{n_x} \times \Nb_{n_\xi} \} \in \Rb^{n_x
%   \times n_\xi}, \qquad t \in [0,T]
% \]
% and the lexicographical representation of the same vector, obtained by ordering
% gridpoints $\{ (x_i,\xi_j) \}$
% and components $\{V_{ij}(t)\}$ via a lexicographical mapping $k(i,j)$,
% \begin{equation}\label{eq:uVec}
%   U(t) = \{ U_{k(i,j)} (t) = V_{k(i,j)}(t) \colon  
%   (i,j) \in \Nb_{n_x} \times \Nb_{n_\xi} \}
%   \in \Rb^{n_xn_\xi}, \qquad t \in [0,T]
%   .
% \end{equation}
%
Collocating \eqref{eq:systemNum} at the somato-dendritic nodes we obtain
\begin{equation}\label{eq:collocation}
  \begin{split}
    \partial_t V(x_j,\xi_i,t) = (-\gamma + \nu \partial_{\xi\xi}) V(x_j,\xi_i,t) 
   & + K(V)(x_j, \xi_i,t) \\
   & + G(x_j,\xi_i,t) \quad
  (j,i) \in \Nb_{n_x} \times \Nb_{n_\xi},
  \end{split}
\end{equation}
where, with a slight abuse of notation, we denote by $V$ an interpolant to the
function $V$ in \eqref{eq:systemNum} through $\{ (x_j,\xi_i) \}$.
%
To obtain a numerical solution of the problem we
must choose: (i) an approximation for the linear operator $(-\gamma + \nu
\partial_{\xi\xi})$ at the somato-dendritic nodes; (ii) an approximation for the
integral operator at the same nodes; (iii) a scheme to time step the derived set of
ODEs.

In the presentation of the scheme, we shall use two equivalent representations for
the voltage approximation: a matricial description 
\begin{equation}\label{eq:VMatrix}
  V(t) = \{ V_{ij}(t) 
  \colon (i,j) \in \Nb_{n_\xi} \times \Nb_{n_x} \} 
    \in \Rb^{n_\xi \times n_x}, \qquad V_{ij}(t) \approx
  V(x_j,\xi_i,t), 
\end{equation}
and a lexicographic vectorial representation, obtained by introducing the
index mapping $k(i,j) = n_\xi(i-1) + j$,
\begin{equation}\label{eq:VVector}
  U(t) = \{ U_{k(i,j)}(t) \colon (i,j) \in \Nb_{n_\xi} \times \Nb_{n_x} \} 
  \in \Rb^{n_x n_\xi}.
\end{equation}
In the latter, we will sometimes suppress the dependence of $k$ on $(i,j)$, for
notational convenience.

\subsection{Discretisation of the linear operator} A simple choice for discretising
the linear differential operator $(-\gamma + \nu \partial_{\xi \xi})$ is to adopt
differentiation matrices~\cite{trefethen2000}. If a differentiation matrix
$D_{\xi \xi} \in \Rb^{n_\xi \times n_\xi}$ is chosen to approximate the action of
the Laplacian operator $\partial_{\xi \xi}$ on twice differentiable, univariate
functions defined on $[-L_\xi,L_\xi]$, satisfying Neumann boundary conditions, and
sampled at nodes $\{ \xi_i \}$, then the action of the operator $-\gamma + \nu
\partial_{\xi\xi}$ on bivariate functions defined on $[-L_x,L_x) \times
[-L_\xi,L_\xi]$, twice differentiable in $\xi$ with Neumann boundary conditions,
sampled at the nodes $\{ (x_j,\xi_i) \}$ with lexicographical ordering $k(i,j)$ is
approximated by the following block-diagonal matrix
\begin{equation*}\label{eq:LinOp}
  -\gamma I_{n_x n_\xi} + \nu I_{n_x} \otimes D_{\xi \xi} = 
  \begin{bmatrix}
    -\gamma + \nu D_{\xi \xi} &                          &        &               \\
                             &-\gamma + \nu D_{\xi \xi} &        &               \\
                             &                          & \ddots &               \\
                             &                          &        &-\gamma + \nu D_{\xi \xi}     \end{bmatrix}
  ,
\end{equation*}
where $I_n$, $n \in \Nb$, is the $n$-by-$n$ identity matrix, and $\otimes$ is the
Kronecker product between matrices. Since the model prescribes diffusion only along
the dendritic coordinate, the
corresponding matrix has a block-diagonal structure \emph{with identical blocks},
which can be exploited to improve performance in numerical computations. The sparsity pattern of a block is dictated by the
underlying scheme to approximate the univariate Laplacian: we have full blocks if
$D_{\xi\xi}$ is derived from spectral schemes, and sparse blocks
for finite-difference schemes.

\subsection{Discretisation of the nonlinear integral operator} The starting point to
discretise the integral operator is an $m$th order quadrature formula with $q_m$
nodes $\{ (y_l,\eta_l) \colon l \in \Nb_{q_m} \}$ and weights $\{ \sigma_l \colon l
\in \Nb_{q_m} \}$ for the integral of a bivariate function over $\Omega$,
\[
Q(v)=\int_{\Omega} v(y,\eta) \, \diff y  \diff \eta \approx
\sum_{l \in \Nb_m} v(y_l,\eta_l) \sigma_l = Q_m(v).
\]
Using this formula we approximate the nonlinear operator in \eqref{eq:collocation} by
\[
  Q_m(K(V))(x_j,\xi_i,t) = \sum_{l \in \Nb{q_m}} W(x_j,\xi_i,y_l,\eta_l)
  S(V(y_l,\eta_l,t)) \sigma_l .
\]

We stress that, in general, the quadrature nodes $\{ (y_l,\eta_l) \}$ and the
collocation nodes $\{ (x_{k(i,j)}, \xi_{k(i,j)}) \}$ are disjoint. The former are
chosen so as to approximate accurately the integral term, the latter to approximate
the differential operator. When the two grids are disjoint, an interpolation of $V$ with
nodes $\{ (y_l,\eta_l) \}$ is necessary to derive a set of ODEs at the collocation
nodes. In the remainder of this paper we will assume that collocation and quadrature
nodes coincide, so that we can omit the interpolant, for simplicity.

\subsection{Matrix ODE formulation}
Combining the differentiation matrix, the quadrature rule, and the lexicographic
representation \eqref{eq:VVector}
we obtain a set of $n_x n_\xi$ ODEs
\begin{equation}\label{eq:ODEs}
  \begin{aligned}
    \dot U(t) & = (-\gamma I_{n_x n_\xi} + \nu I_{n_x} \otimes D_{\xi \xi} ) U(t) +
  F(U(t),t), \\
  U(0) & = U_0.
  \end{aligned}
\end{equation}
The structure of the differentiation matrix in section \eqref{eq:LinOp}, however,
suggests a rewriting of \eqref{eq:ODEs} in terms of the blocks of the linear
operator, which correspond to ``slices" at constant values of $x$: we recall the
matrix representation \eqref{eq:VMatrix} and obtain an equivalent
matrix ODE formulation
\begin{equation}\label{eq:MatrixODE}
  \dot V(t) = (-\gamma I_{n_\xi} + \nu D_{\xi \xi}) V(t) + N(V(t)) + G(t),
\end{equation}
where $N$ is the matrix-valued function with components $N_{ij}(V) =
Q_m(V)(x_j,\xi_i)$ and $G$ is the matrix with components $G(x_j,\xi_i,t)$. In passing,
we note that the linear part of the equation involves
a multiplication between an $n_\xi$-by-$n_\xi$ matrix and the $n_\xi$-by-$n_x$ matrix $V$.

\subsection{Time-stepping scheme} The proposed time-stepping scheme for
\eqref{eq:systemNum} is obtained from \eqref{eq:MatrixODE} with the following
choices: (i) a first-order, implicit-explicit (IMEX) time-stepping
scheme~\cite{ascher1995}; (ii) a second-order, centered finite-difference
scheme for the differentiation matrix $D_{\xi \xi}$; (ii) a second-order trapezium
rule for the quadrature rule $Q_m$. As we shall see, these choices bring
a few computational advantages, which will be outlined below.

IMEX schemes treat the linear (diffusive) part of the ODE implicitly, and the
nonlinear part explicitly, so that the stiff diffusive term is integrated implicitly
to avoid excessively small time steps. The simplest IMEX method uses backward Euler
for the diffusive term, leading to
\begin{equation}\label{eq:IMEX}
  \begin{aligned}
       V^0 & = V_0, \\
       A V^n & = V^{n-1} + \tau N(V^{n-1}) + \tau G^{n-1}, \qquad n \in \Nb,
  \end{aligned}
\end{equation}
where $V^n \approx V(t_n)$, $G^n = G(t_n)$, and $A$ is the matrix
\begin{equation}\label{eq:AMatr}
  A = (1+\gamma \tau) I_{n_\xi} - \tau \nu D_{\xi \xi}.
\end{equation}
In concrete calculations we use second-order centred finite differences, leading to
\begin{equation}\label{eq:FinDiffLapl}
  D_{\xi \xi} = 
   \Delta/h_\xi^{2}, \qquad \Delta = 
  \begin{bmatrix}
    -2 &  2     &         &        &    &     \\
     1 & -2     &       1 &        &    &     \\
       & \ddots & \ddots & \ddots  &    &     \\
       &        &      1 &      -2 & 1  &     \\
       &        &        &       2 & -2 &  
   \end{bmatrix},
 \end{equation}
in which Neumann boundary conditions are included in the differentiation matrix. 
% This
% leads to $A$ being tridiagonal, and strictly diagonally dominant \da{Decide what to
% say about positive-definiteness}. \da{Pedro, please amend your eigenvalue
%   computations/spectral radii to the new matrix, which has Neuamnn BCs. The substance
% does not change, but the eigenvalues have a slightly different expression with
% Neumann BCs.}

Finally, we discuss the choice of the quadrature scheme. We use a composite trapezium
scheme with nodes $\{x_j\}$ and weights $\{\rho_j\}$ in $x$, and nodes $\{\xi_i\}$
and weights $\{ \sigma_i\}$ in $\xi$, respectively, hence quadrature and collocation
sets coincide,
\begin{equation} \label{eq:NQuad}
  N_{ij}(V) = \sum_{j' \in \Nb{n_x}} \sum_{i' \in \Nb{n_\xi}}
  W(x_j,\xi_i,x_{j'},\xi_{i'}) S(V_{i',j'}) \rho_{j'}\sigma_{i'}.
  \quad (i,j) \in \Nb_{n_\xi} \times \Nb_{n_x}.
\end{equation}

\subsection{Properties of the IMEX scheme}
In this section we collect some analytical results on the IMEX scheme
\eqref{eq:IMEX}--\eqref{eq:NQuad}. We work with spaces of sufficiently regular
continuous functions, which provides the simplest setting for our results. We denote
by $C^k(D)$ the space of $k$-times continuously differentiable functions from $D$ to
$\Rb$, where $k$ is an integer, $D$ a domain in $\Rb^3$. We also indicate
by $C_b^k(D)$ the space of continuous functions from $D$ to $\Rb$ with bounded and
continuous partial derivatives up to order $k$. 
Both spaces are endowed with the infinity norm $\Vert \blank
\Vert_\infty$. We will use the symbol $| \blank |_\infty$ for the standard
infinity-norm on matrices, induced by the corresponding vector norm. In addition, we
will denote by $\bar D$ the closure of $D$.

We begin with a generic assumption of boundedness on the functions in
\eqref{eq:systemNum}:
\begin{hypothesis}\label{hyp:boundedness}
  There exist $C_W, C_S, C_G >0 $ such that
  \[
    |W| \leq C_W \; \textrm{in $\Omega \times \Omega$},
    \qquad
    |S| \leq C_S \; \textrm{in $\Rb$},
    \qquad
    |G| \leq C_G \; \textrm{in $\Omega \times \Rb$}.
  \]
\end{hypothesis}

\begin{lemma}[Boundedness of IMEX solution]\label{lem:IMEXboundedness} 
  Assume Hypothesis \ref{hyp:boundedness}, then there exists a unique bounded
  sequence $(V^n)_{n\in\Nb}$ satisfying
  the IMEX scheme \eqref{eq:IMEX}--\eqref{eq:NQuad}. In
  addition, the following bound holds
  \[
    \vert V^n \vert_\infty \leq \vert V^0 \vert_\infty + 
    n_x \frac{ \mu(\bar\Omega) C_W C_S + C_G}{\gamma},
    \qquad n \in \Nb.
  \]
\end{lemma}
\begin{proof}
  The matrix $A$ in \eqref{eq:AMatr} has real, strictly positive eigenvalues given
  by
  \[
    \lambda_k = 1 + \gamma \tau + \frac{4\nu \tau}{h_\xi^2} 
    \bigg[ \sin \bigg( \frac{\pi(k-1)}{2n_\xi} \bigg) \bigg]^2, \qquad k \in
    \Nb_{n_\xi},
  \]
  where we have used the fact that the eigenvalues of $D_{\xi \xi}$ are known in closed form.
  We conclude that $A$ is invertible, hence for any fixed $n \in \Nb$,
  the matrix $V^n$ solving \eqref{eq:IMEX} is unique. In addition, $A$ is
  strictly diagonally dominant, hence the following bound holds~\cite{varah75}
  \begin{equation}\label{eq:invABound}
  \vert A^{-1} \vert_\infty \leq \max_{i \in \Nb_\xi} \frac{1}{|A_{ii}| - \sum_{j
  \neq i} |A_{ij}|} = \frac{1}{1+ \gamma \tau}.
  \end{equation}
  To prove boundedness of the sequence $(V^n)_{n \in \Nb}$ we first bound the
  matrices $N(V^{n-1})$, $G^n$ appearing in \eqref{eq:IMEX}
  \[
    \begin{aligned}
    \vert N(V^{n-1}) \vert_\infty 
    & = \max_{i \in \Nb_{n_\xi}} \sum_{j \in \Nb_{n_x}} |N_{ij}(V^{n-1})| \\
    & \leq
    \max_{i \in \Nb_{n_\xi}} \sum_{j \in \Nb_{n_x}} 
                             \sum_{j' \in \Nb{n_x}} \sum_{i' \in \Nb{n_\xi}}
                              |W(x_j,\xi_i,x_{j'},\xi_{i'}) S(V_{i',j'}) \rho_{j'}\sigma_{i'}| \\
    & \leq C_W C_S
    \max_{i \in \Nb_{n_\xi}} \sum_{j \in \Nb_{n_x}} 
                   \sum_{j' \in \Nb{n_x}} \sum_{i' \in \Nb{n_\xi}}
		  \rho_{j'}\sigma_{i'} \\
    & \leq C_W C_S
    \max_{i \in \Nb_{n_\xi}} \sum_{j \in \Nb_{n_x}} \mu(\bar\Omega)
    = n_x \mu(\bar\Omega) C_W C_S,
    \end{aligned},
  \]
  and similarly $\vert G^{n-1} \vert_\infty \leq n_x C_G$, and then combine
  them with the bound for $\vert A^{-1}\vert_\infty$ to find
  \[
    \begin{aligned}
    \vert V^n \vert_\infty 
        & \leq \vert A^{-1} \vert_\infty 
	  \Big(
	      \vert V^{n-1} \vert_\infty 
	    + \tau \vert N(V^{n-1}) \vert_\infty
	    + \tau \vert G^{n-1} \vert_\infty 
	  \Big) \\
% 	& \leq \Vert A^{-1} \Vert_\infty 
% 	  \Big(
% 	      \Vert V_i^{n-1} \Vert_\infty 
% 	      + \tau n_x |\Omega| \Vert w \Vert_\infty \Vert S \Vert_\infty 
% 	      + \tau n_x \Vert G \Vert_\infty
% 	  \Big) \\
	& \leq \frac{1}{1+\gamma \tau}
	  \Big(
	      \vert V^{n-1} \vert_\infty 
	      + \tau n_x \mu(\bar\Omega) C_W C_S 
	      + \tau n_x C_G
	  \Big).
    \end{aligned}
    % \quad (i,n) \in \Nb_{n_x} \times \Nb
  \]
  We set
  \[
    r = \frac{1}{1+\gamma \tau} <1, \qquad 
    q = \frac{\tau n_x}{1 + \gamma \tau}
      ( \mu(\bar\Omega) C_W C_S + C_G),
  \]
  and use induction and elementary properties of the geometric series to obtain
  \[
    \vert V^n \vert_\infty \leq r^n \vert V^0 \vert_\infty + q \sum_{j =0}^{n-1} r^j
      \leq \vert V^0 \vert_\infty + \frac{q}{1-r},
  \]
  which proves the assertion.
\end{proof}

In addition to proving boundedness of the solution, we address the
convergence rate of the IMEX scheme. For this result, we assume the existence of a
sufficiently regular solution to \eqref{eq:systemNum}.

\begin{lemma}[Local convergence rate of the IMEX scheme]\label{lem:IMEXconvergence}
  Assume Hypothesis \ref{hyp:boundedness}, $W \in C^2(\Omega \times \Omega)$, $S
  \in C^2_b(\Omega)$, and assume \eqref{eq:systemNum}
  admits a strong solution $V_*$ 
  whose partial derivatives $\partial_{tt}V_*$, $\partial_{xx}V_*$,
